Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-16 application.tex
revised

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume temperature fluctuations are smoothed out
by radiative diffusion and model this in linear theory by writing the
linearized RHS of Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%   c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
% where $\hat{\eta} = \eta/H^2\Omega_k$ is the 
% dimensionless diffusion coefficient.

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

@@

%but
%given that the VSI depends on vertical shear, an effect that becomes
%important away from the disk midplane, the inner disk may still be
%able to develop the VSI, but it is likely restricted to the disk
%surface.   
%We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
%at small radii by choosing a sufficiently large $\khat$. However, our
%linear calculations suggest growth rates eventually decay as 
%$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
%probably unimportant.

\begin{figure*}
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z3}\\ 
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  1cm]{figures/bcrit_mmsn_kap1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm
  1.8cm 0cm 1cm]{figures/bcrit_mmsn_kap1_z3}\\
  \includegraphics[scale=.47,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z3} 
  \caption{Dimensionless thermal relaxation timescales $\beta$,
    evaluated at the midplane (left) and at $z=3H$ (right) in the
    fiducial protoplanetary disk. Eq. \ref{beta_mmsn} is plotted  
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the 
    critical thermal relaxation timescale $\beta_\mathrm{crit}$.  
    % and 
    % $\beta<\beta_\mathrm{crit}$ is necessary for the fundamental isothermal VSI to operate  
    % for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure*}

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at 
each radius the value of $\khat$ such  
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI would only
occur on very small radial scales in the inner disk. % However, these
% may not be important as 
% our
% linear calculations suggest growth rates eventually decay for large
% $\khat$.

@@

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height.
Accordingly we expect, and find, the VSI operates at $r=50$AU. The
inner disk ($r=5$AU) is optically thick near the midplane but becomes
optically thin away from it, such that $\beta < \beta_\mathrm{crit}$
for $|z|\gtrsim1.25H$. In fact, we also find the VSI at $r=5$AU, but
with much reduced growth rates (when scaled by the local rotation).

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$.  In fact, we either only find the fundamental mode, or
that it is the most unstable, except for the appearence of some
surface modes towards the inner disk when considering large
wavenumbers $\khat\gtrsim 50$. However,
surface modes are affected by boundary conditions and disturbances
with large $\khat$ are more likely subject to viscous decay, as
discussed above. The dominance of the fundamental mode is also
observed in hydrodynamic simulations of \cite{stoll14} which treat
radiative cooling properly.

@@

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane. %because of fundamental mode

>>>>

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume temperature fluctuations are smoothed out
by radiative diffusion and model this in linear theory by writing the
linearized RHS of Eq. \ref{real_energy} as

\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
fluctuations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

@@

\begin{figure*}
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z3}\\ 
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  1cm]{figures/bcrit_mmsn_kap1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm
  1.8cm 0cm 1cm]{figures/bcrit_mmsn_kap1_z3}\\
  \includegraphics[scale=.47,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z3} 
  \caption{Dimensionless thermal relaxation timescales $\beta$,
    evaluated at the midplane (left) and at $z=3H$ (right) in the
    fiducial protoplanetary disk. Eq. \ref{beta_mmsn} is plotted  
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the 
    critical thermal relaxation timescale $\beta_\mathrm{crit}$.  
    \label{mmsn_bcrit_bcool}}   
\end{figure*}

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at 
each radius the value of $\khat$ such  
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI would only
occur on very small radial scales in the inner disk.

@@

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height.
Accordingly we expect, and find, the VSI operates at $r=50$AU. The
inner disk ($r=5$AU) is optically thick near the midplane but becomes
optically thin away from it, such that $\beta < \beta_\mathrm{crit}$
for $|z|\gtrsim1.25H$. In fact, we also find the VSI in this case, but
with much reduced growth rates (when scaled by the local rotation), as
described below.

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$.  In fact, we either only find the fundamental mode, or
that it is the most unstable, except for the appearance of some
surface modes towards the inner disk when considering large
wavenumbers $\khat\gtrsim 50$. However,
surface modes are affected by boundary conditions and disturbances
with large $\khat$ are more likely subject to viscous decay, as
discussed above. The dominance of the fundamental mode is also
observed in hydrodynamic simulations of \cite{stoll14} which treat
radiative cooling properly.

@@

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

### 2015-03-16 linear_adia.tex
about to edit numerical
revised

We insert Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso_explicit}, keeping only first order terms, and
solve for $\delta\hat{\sigma}$ using the above xpressions for $\delta\chi$
and $\hat{\sigma}$. We find imaginary part of $\delta\hat{\sigma}$
is 
% yield the requirements, after
% balancing constants and coefficients of $\zhat^2$, 
% \begin{align}
%   0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
%   \left(\hat{\sigma}^2 -1\right)\delta\chi, \\
%   0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
%     q \khat\right)\delta\chi.
% \end{align}
% This gives 
% \begin{align}\label{nearly_iso_dsig}
%   \delta \hat{\sigma} =
%   -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
%       q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
% \end{align}
% The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization, since $\gamma>1$. 
% Interestingly, for both
% $\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
% stabilization by small finite cooling is ineffective at very large or
% very small scales.

@@

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$ Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since the second term in square brackets is small by assumption, we have 
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without assuming $|\epsilon q\khat|\lesssim 1$.  We will confirm Eq. \ref{iso_vsi_cond} numerically in 
\S\ref{bcrit_num_test}.

>>>>

We insert Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso_explicit}, keeping only first order terms, and
solve for $\delta\hat{\sigma}$ using the above expressions for $\delta\chi$
and $\hat{\sigma}$. We find imaginary part of $\delta\hat{\sigma}$
is 
% yield the requirements, after
% balancing constants and coefficients of $\zhat^2$, 
% \begin{align}
%   0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
%   \left(\hat{\sigma}^2 -1\right)\delta\chi, \\
%   0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
%     q \khat\right)\delta\chi.
% \end{align}
% This gives 
% \begin{align}\label{nearly_iso_dsig}
%   \delta \hat{\sigma} =
%   -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
%       q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
% \end{align}
% The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization, since $\gamma>1$. 
% Interestingly, for both
% $\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
% stabilization by small finite cooling is ineffective at very large or
% very small scales.

@@

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|\epsilon q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without assuming $|\epsilon q\khat|\lesssim 1$.
We will confirm $\beta_\mathrm{crit}$ numerically in 
\S\ref{bcrit_num_test}.
